---
title: "Introduction to CBAMMR: Comprehensive Meta-Analysis with 2025 Standards"
author: "CBAMMR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CBAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

CBAMMR (Comprehensive Bayesian and Advanced Meta-Analysis Methods in R) is a cutting-edge R package that implements state-of-the-art meta-analysis methods following 2024-2025 journal standards including PRISMA 2020, GRADE, and advanced clinical decision-making tools.

### Key Features

- **Comprehensive Methods**: 60+ functions covering all modern meta-analysis approaches
- **Transportability Weighting**: Adjust results for target populations using entropy balancing
- **GRADE Integration**: Auto-assessment of evidence certainty
- **Clinical Decision Support**: Fragility index, NNT by risk, decision curves, MID assessment
- **Reproducibility**: Complete environment capture and OSF/Zenodo export bundles
- **Reporting Standards**: PRISMA 2020 compliance, power analysis
- **Publication-Ready**: Tables, plots, and outputs for top-tier journals

## Installation

```{r eval=FALSE}
# From GitHub
devtools::install_github("mahmood726-cyber/CBAMMR")

# Load package
library(CBAMMR)
```

## Quick Start: Binary Outcome Meta-Analysis

This example demonstrates a complete workflow from data to publication-ready outputs.

### Step 1: Prepare Data

```{r eval=FALSE}
library(CBAMMR)

# Simulate example data (or load your own)
set.seed(123)
data <- simulate_cbamm_binary(
  n = 20,                    # 20 studies
  measure = "OR",            # Odds ratio
  true_effect = 0.70         # True OR = 0.70 (benefit)
)

# Add study characteristics for transportability
data$age_mean <- rnorm(20, 60, 10)
data$female_pct <- runif(20, 0.3, 0.7)
data$study_type <- sample(c("RCT", "Obs"), 20, replace = TRUE, prob = c(0.6, 0.4))

# View data structure
head(data[, c("study", "ai", "bi", "ci", "di", "age_mean", "female_pct")])
```

### Step 2: Configure Analysis

```{r eval=FALSE}
# Set up analysis configuration
config <- setup_cbamm(
  effect_measure = "OR",
  use_transport = TRUE,      # Transportability weighting
  use_hksj = TRUE,          # Hartung-Knapp adjustment
  use_bayesian = FALSE,     # Skip for quick example
  run_mv = TRUE,            # Multivariate meta-analysis
  export_results = TRUE     # Save all outputs
)
```

### Step 3: Define Target Population

```{r eval=FALSE}
# Define characteristics of your target population
target_pop <- list(
  age_mean = 65,       # Average age of target population
  female_pct = 0.52    # Proportion female in target population
)
```

### Step 4: Run Complete Analysis

```{r eval=FALSE}
# Run comprehensive meta-analysis
results <- run_cbamm_analysis(
  data = data,
  target_population = target_pop,
  config = config
)

# The analysis runs:
# 1. Transportability weighting (if target_pop provided)
# 2. GRADE quality down-weighting
# 3. Random-effects meta-analysis (REML with HKSJ)
# 4. Stratified analysis (RCT vs Obs)
# 5. Publication bias tests (Egger, PET-PEESE)
# 6. Influence analysis
# 7. Multivariate meta-analysis
# 8. All visualizations
# 9. Summary tables
```

### Step 5: View Main Results

```{r eval=FALSE}
# Main pooled effect (transportability-weighted)
print(results$pooled$transport)

# Access specific results:
# - Pooled effect: HR 0.70 [95% CI: 0.60, 0.82]
# - Heterogeneity: I² = 45%, τ² = 0.04
# - Prediction interval: [0.50, 0.98]
```

## Advanced Features: 2025 Journal Standards

### PRISMA 2020 Compliance

Generate a complete PRISMA 2020 checklist for your manuscript:

```{r eval=FALSE}
# Generate PRISMA checklist
prisma <- cbamm_prisma_checklist(
  populated = TRUE,
  results = results
)

# Save for manuscript
readr::write_csv(prisma, "prisma_checklist.csv")

# All 27 items with auto-population where possible
View(prisma)
```

### GRADE Evidence Assessment

Assess certainty of evidence with auto-assessment:

```{r eval=FALSE}
# Generate GRADE profile
grade <- cbamm_grade_profile(
  results = results,
  data = data,
  outcome_name = "All-cause mortality"
)

# View certainty rating
print(grade$final_certainty)
# e.g., "MODERATE" (started HIGH, downgraded for inconsistency)

# View domain assessments
print(grade$profile)
#   Domain          Assessment      Explanation
#   Risk of Bias    not serious     Based on study design
#   Inconsistency   serious         I² = 45%
#   Imprecision     not serious     CI width acceptable
#   ...

# Save for manuscript
readr::write_csv(grade$profile, "grade_assessment.csv")
```

### Fragility Index (Statistical Robustness)

Assess how robust your findings are to small changes:

```{r eval=FALSE}
# Calculate fragility index
fragility <- cbamm_fragility_index(
  results = results,
  data = data
)

# Interpret
print(fragility$interpretation)
# "ROBUST: FI = 28. Result is robust and precise"
# (FI ≥ 22 = highly robust; FI ≤ 5 = fragile)

# Include in Discussion section:
# "The fragility index was 28, indicating statistically robust findings."
```

### NNT Stratified by Patient Risk

Show how absolute benefit varies by baseline risk:

```{r eval=FALSE}
# Calculate NNT across risk levels
nnt_table <- cbamm_nnt_by_baseline_risk(
  results = results,
  baseline_risks = c(0.01, 0.05, 0.10, 0.20, 0.40)
)

# Results show:
#   baseline_risk   nnt   direction
#   0.01            500   NNTB     (low-risk patients)
#   0.05            100   NNTB
#   0.10             50   NNTB
#   0.20             25   NNTB
#   0.40             12   NNTB     (high-risk patients)

# Interpretation: High-risk patients benefit much more than low-risk
```

### Clinical Significance (Beyond P-values)

Assess whether effect exceeds minimal important difference:

```{r eval=FALSE}
# Define MID threshold
# For OR: 0.10 means OR must be < 0.90 or > 1.10 to be clinically meaningful
clinical_sig <- cbamm_clinical_significance(
  results = results,
  mid = 0.10,
  mid_source = "Established clinical guidelines"
)

# View classification
print(clinical_sig$classification)
# "Clinically significant (robust)" - effect AND CI exceed MID
# "Clinically significant (point estimate only)" - only effect exceeds MID
# "Not clinically significant" - effect < MID threshold
```

### Prediction Interval Interpretation

Assess whether treatment will work in different settings:

```{r eval=FALSE}
# Interpret PI with clinical thresholds
pi_threshold <- cbamm_prediction_interval_threshold(
  results = results,
  benefit_threshold = 0.80,  # OR < 0.80 = clinically beneficial
  harm_threshold = 1.20       # OR > 1.20 = harmful
)

# Results:
print(pi_threshold$interpretation)
# "High confidence: PI entirely in beneficial range"
# Probability future study shows benefit: 92%

# Include in Discussion:
# "The 95% prediction interval [0.50, 0.98] suggests high confidence
#  that the intervention will be beneficial in most clinical settings."
```

### Decision Curve Analysis (Net Clinical Benefit)

Determine optimal treatment strategy across patient preferences:

```{r eval=FALSE}
# Calculate net benefit
net_benefit <- cbamm_net_clinical_benefit(
  results = results,
  data = data,
  baseline_risk = 0.15
)

# View decision curve
plot(net_benefit$decision_curve)
# Shows when intervention provides net benefit vs treat-all or treat-none

# Optimal range
print(net_benefit$optimal_threshold_range)
# Intervention optimal for threshold probabilities: 0.10 - 0.60
```

### Power Analysis

Calculate power or required sample size:

```{r eval=FALSE}
# What power do we have?
power_achieved <- cbamm_power_analysis(
  effect_size = results$pooled$transport$beta[1],
  tau2 = results$pooled$transport$tau2,
  n_studies = nrow(data),
  avg_n_per_study = 100
)
print(power_achieved$power)  # e.g., 0.85 (85% power)

# How many studies needed for 90% power?
n_required <- cbamm_power_analysis(
  effect_size = 0.3,
  tau2 = 0.04,
  power = 0.90,
  avg_n_per_study = 100
)
print(n_required$n_studies_required)  # e.g., 32 studies
```

## Complete Manuscript Workflow

This example shows a full workflow from analysis to publication:

```{r eval=FALSE}
# 1. Run comprehensive analysis
results <- run_cbamm_analysis(data, target_pop, config)

# 2. Generate all reporting outputs
prisma <- cbamm_prisma_checklist(populated = TRUE, results = results)
grade <- cbamm_grade_profile(results, data, outcome_name = "Mortality")
power <- cbamm_power_analysis(0.3, 0.04, n_studies = 20)

# 3. Clinical decision-making analyses
fragility <- cbamm_fragility_index(results, data)
nnt_table <- cbamm_nnt_by_baseline_risk(results, c(0.05, 0.10, 0.20, 0.40))
clinical_sig <- cbamm_clinical_significance(results, mid = 0.10)
pi_threshold <- cbamm_prediction_interval_threshold(results)
net_benefit <- cbamm_net_clinical_benefit(results, data, baseline_risk = 0.15)

# 4. Create reproducibility bundle for supplements
cbamm_export_bundle(
  results = results,
  data = data,
  config = config,
  output_dir = "supplementary_materials",
  include_data = TRUE
)

# 5. Results folder now contains:
# - supplementary_materials/
#   ├── data/analysis_data.csv
#   ├── code/reproduce_analysis.R
#   ├── results/*.csv
#   ├── documentation/codebook.csv
#   └── README.md

# 6. Upload to OSF/Zenodo, get DOI, cite in manuscript
```

## Manuscript Text Examples

### Methods Section

```
"Meta-analysis was conducted using CBAMMR v7.0 (https://github.com/mahmood726-cyber/CBAMMR)
in R version 4.4.0. We used random-effects meta-analysis with restricted maximum likelihood
(REML) estimation and Hartung-Knapp-Sidik-Jonkman adjustments. Transportability weights were
applied to adjust estimates for our target population (mean age 65, 52% female) using entropy
balancing. Certainty of evidence was assessed using GRADE. Statistical robustness was evaluated
using the fragility index. Clinical significance was assessed relative to a minimal important
difference of 0.10 on the odds ratio scale. Number needed to treat was calculated across baseline
risk levels from 1% to 40%. All analyses followed PRISMA 2020 guidelines."
```

### Results Section

```
"Twenty studies (n = 4,500 participants) were included. The pooled odds ratio was 0.70
(95% CI: 0.60, 0.82; I² = 45%, prediction interval: 0.50-0.98), indicating moderate
heterogeneity. The effect exceeded our pre-specified minimal important difference threshold,
suggesting clinical significance. The fragility index was 28, indicating robust findings.
Number needed to treat varied from 500 in low-risk patients (1% baseline risk) to 12 in
high-risk patients (40% baseline risk). GRADE assessment yielded MODERATE certainty evidence
(downgraded for inconsistency). Decision curve analysis indicated net clinical benefit for
threshold probabilities between 0.10 and 0.60."
```

### Discussion Section

```
"Our meta-analysis achieved 85% power to detect the observed effect. The prediction interval
suggests high confidence that future studies will demonstrate benefit in most settings.
Statistical robustness was confirmed by a fragility index of 28, well above the threshold
of 22 for highly robust findings. The substantial variation in NNT by baseline risk
highlights the importance of risk stratification in clinical decision-making."
```

## Continuous Outcomes Example

For continuous outcomes (e.g., standardized mean difference):

```{r eval=FALSE}
# Simulate continuous data
data_cont <- simulate_cbamm_continuous(
  n = 18,
  measure = "SMD",
  true_effect = -0.40    # SMD = -0.40 (moderate benefit)
)

# Configure
config_cont <- setup_cbamm(
  effect_measure = "SMD",
  use_hksj = TRUE
)

# Run analysis
results_cont <- run_cbamm_analysis(
  data = data_cont,
  target_population = NULL,  # No transportability for this example
  config = config_cont
)

# Assess clinical significance using MID
# For SMD, Cohen's d = 0.20 is "small", 0.50 "medium", 0.80 "large"
clinical_sig_cont <- cbamm_clinical_significance(
  results = results_cont,
  mid = 0.20,  # Small effect threshold
  measure = "SMD",
  mid_source = "Cohen's conventions"
)
```

## Visualization

CBAMMR automatically generates publication-ready plots:

```{r eval=FALSE}
# All plots saved to cbamm_outputs/
# - Forest plot with study details
# - Funnel plot for publication bias
# - Leave-one-out sensitivity plot
# - Cumulative meta-analysis plot
# - Multiverse analysis (specification curve)
# - Prediction interval visualization
# - And more...

# Display all plots interactively
cbamm_show_all_plots(results)

# Or save as PDF
cbamm_show_all_plots(
  results,
  save_dir = "figures",
  pdf_file = "meta_analysis_plots.pdf",
  width = 10,
  height = 8
)
```

## Summary Tables

Export publication-ready summary tables:

```{r eval=FALSE}
# Comprehensive results table
summary_table <- cbamm_make_summary_table(
  results = results,
  data = data,
  config = config
)

# Includes:
# - Pooled effects (multiple methods)
# - Heterogeneity statistics
# - Publication bias tests
# - Rare events methods (if applicable)
# - Bayesian results (if run)
# - E-values
# - And more...

# Save for manuscript
readr::write_csv(summary_table, "results_table.csv")
```

## Tips for Best Results

### 1. Data Preparation
- Ensure effect size columns (yi, sei) OR raw data (ai, bi, ci, di) are present
- Include study identifiers, year, sample sizes
- Add study characteristics (age, sex, quality) for subgroup/meta-regression

### 2. Configuration
- Use `use_hksj = TRUE` for better CI coverage
- Enable transportability if target population differs from study populations
- Consider multivariate models for correlated outcomes

### 3. Reporting
- Always generate PRISMA checklist and GRADE profile
- Include fragility index in Discussion
- Present NNT stratified by risk for binary outcomes
- Report prediction intervals, not just confidence intervals

### 4. Reproducibility
- Always export bundle for supplementary materials
- Include exact package versions
- Share data and code on OSF/Zenodo

## Package Philosophy

CBAMMR is designed with three core principles:

1. **Comprehensiveness**: Implements all modern meta-analysis methods in one place
2. **Reproducibility**: Complete environment tracking and easy sharing
3. **Journal-Readiness**: Meets 2025 standards from PRISMA, GRADE, and top journals

## Citation

When using CBAMMR in publications, please cite:

```
CBAMMR v7.0: Comprehensive Bayesian and Advanced Meta-Analysis Methods in R
https://github.com/mahmood726-cyber/CBAMMR
```

And cite the reporting standards you followed:

- PRISMA 2020: Page MJ, et al. (2021). BMJ, 372:n71
- GRADE: Guyatt GH, et al. (2011). J Clin Epidemiol, 64(4):383-394

## Getting Help

- Documentation: `?function_name` (e.g., `?cbamm_grade_profile`)
- Examples: `example(run_cbamm_analysis)`
- GitHub Issues: https://github.com/mahmood726-cyber/CBAMMR/issues

## Additional Resources

- PRISMA 2020: https://www.prisma-statement.org/
- GRADE Handbook: https://gradepro.org/handbook
- Cochrane Handbook: https://training.cochrane.org/handbook
- Doing Meta-Analysis in R: https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/

---

**Package Version**: 7.0.0
**Last Updated**: `r Sys.Date()`
