---
title: "Clinical Decision-Making from Meta-Analysis"
author: "CBAMMR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Decision-Making from Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Meta-analysis traditionally answers: "Does the treatment work on average?" But clinicians and policy-makers need answers to more practical questions:

1. 📊 **"Should we use this treatment?"** (Decision curve analysis)
2. 💰 **"Is more research worth funding?"** (EVPI)
3. 🎯 **"How certain is this recommendation?"** (Threshold analysis)
4. 👤 **"Will it work for MY patient?"** (Individualized effects)
5. 📈 **"What's the absolute benefit?"** (NNT)

CBAMMR v8.1 provides **five clinical decision tools** that translate meta-analytic evidence into actionable insights. These tools are based on recent research in medical decision-making and health economics.

This vignette demonstrates all five methods using a realistic clinical scenario.

## Case Study: Novel Anticoagulant for Stroke Prevention

We'll examine a meta-analysis of a novel oral anticoagulant (NOAC) versus warfarin for stroke prevention in atrial fibrillation. This scenario involves:

- **Real clinical decision:** Should we switch patients to the new drug?
- **Cost considerations:** NOACs are more expensive than warfarin
- **Patient heterogeneity:** Different patients have different stroke/bleeding risks
- **Policy implications:** Should guidelines recommend NOACs?

### The Data

```{r}
library(CBAMMR)
library(metafor)

# Meta-analysis data from 12 RCTs
# Effect size: log(HR) for stroke (negative = NOAC better)
yi <- c(-0.18, -0.22, -0.15, -0.28, -0.19, -0.25, -0.16, -0.23, -0.20, -0.17, -0.26, -0.21)

# Sampling variances
vi <- c(0.012, 0.015, 0.018, 0.011, 0.014, 0.013, 0.016, 0.012, 0.015, 0.017, 0.013, 0.014)

# Additional data for individualized effects
# Patient characteristics: age, prior stroke, CHADS2 score
moderators <- data.frame(
  age = c(68, 72, 65, 74, 70, 69, 71, 67, 73, 66, 75, 68),
  prior_stroke = c(0, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0),
  chads2 = c(2, 3, 2, 4, 2, 3, 3, 2, 4, 2, 4, 3)
)

cat("Number of studies:", length(yi), "\n")
cat("Mean log(HR):", round(mean(yi), 3), "\n")
cat("Mean HR:", round(exp(mean(yi)), 3), "( NOAC reduces stroke by ~20%)\n")
```

## Tool 1: Decision Curve Analysis

**Question:** At what risk threshold should we prescribe NOACs instead of warfarin?

Decision curve analysis finds the optimal decision threshold by maximizing **net benefit**, accounting for both benefits and harms.

### Concept

- **Benefit:** Preventing strokes
- **Harm:** Increased bleeding risk, cost, monitoring burden
- **Trade-off:** Harm/benefit ratio

### Analysis

```{r}
# Decision curve analysis
# Harm/benefit ratio = 0.8 (mild increase in bleeding for major stroke prevention)
dca_result <- cbamm_decision_curve(
  yi = yi,
  vi = vi,
  threshold_range = seq(0.01, 0.10, by = 0.005),
  harm_benefit_ratio = 0.8
)

print(dca_result)
```

### Visualization

```{r, fig.height=5}
plot(dca_result)
```

### Clinical Interpretation

```{r}
cat("\n╔═══════════════════════════════════════════════════════╗\n")
cat("║       DECISION CURVE ANALYSIS RESULTS                 ║\n")
cat("╚═══════════════════════════════════════════════════════╝\n\n")

cat("Optimal threshold:", round(dca_result$optimal_threshold, 3), "\n")
cat("Interpretation: Prescribe NOAC if patient's annual stroke risk > ",
    round(dca_result$optimal_threshold * 100, 1), "%\n\n", sep = "")

cat("CLINICAL GUIDELINES:\n")
cat("─────────────────────\n")
cat("✓ High risk (>5% annual): NOAC recommended\n")
cat("✓ Moderate risk (2-5%):   NOAC preferred\n")
cat("? Low risk (<2%):         Warfarin acceptable\n")
```

## Tool 2: Expected Value of Perfect Information (EVPI)

**Question:** Should we fund additional RCTs, or is current evidence sufficient?

EVPI calculates the **maximum value of eliminating uncertainty** through additional research.

### Concept

- If we're uncertain, we might make the wrong decision
- Additional research reduces uncertainty
- EVPI = Expected cost of making wrong decisions × Probability of being wrong

### Analysis

```{r}
# EVPI calculation
# Annual stroke in AF patients in US: ~100,000
# Preventing one stroke saves ~$50,000
evpi_result <- cbamm_evpi(
  yi = yi,
  vi = vi,
  benefit_per_unit = 50000,  # $ per stroke prevented
  population_size = 100000,  # Annual AF patients
  time_horizon = 10,        # Years
  discount_rate = 0.03      # 3% annual discount
)

print(evpi_result)
```

### Visualization

```{r, fig.height=5}
plot(evpi_result)
```

### Research Funding Decision

```{r}
cat("\n╔═══════════════════════════════════════════════════════╗\n")
cat("║     EVPI: SHOULD WE FUND MORE RESEARCH?               ║\n")
cat("╚═══════════════════════════════════════════════════════╝\n\n")

cat("EVPI per person:      $", round(evpi_result$evpi_per_person, 2), "\n")
cat("EVPI for population:  $", format(round(evpi_result$evpi_total), big.mark = ","), "\n")
cat("Annual EVPI:          $", format(round(evpi_result$evpi_total / evpi_result$time_horizon),
                                      big.mark = ","), "/year\n\n")

# Compare to cost of RCT
rct_cost <- 10000000  # $10 million for large RCT
cat("Cost of additional RCT: $", format(rct_cost, big.mark = ","), "\n\n")

if (evpi_result$evpi_total > rct_cost) {
  cat("RECOMMENDATION: ✓ FUND MORE RESEARCH\n")
  cat("Value of information (", format(round(evpi_result$evpi_total), big.mark = ","),
      ") EXCEEDS research cost\n")
  cat("Expected ROI:", round((evpi_result$evpi_total / rct_cost - 1) * 100, 0), "%\n")
} else {
  cat("RECOMMENDATION: ✗ DO NOT FUND MORE RESEARCH\n")
  cat("Current evidence is sufficient. Focus resources elsewhere.\n")
}
```

## Tool 3: Threshold Analysis for Decision Robustness

**Question:** How confident should we be in the guideline recommendation?

Threshold analysis quantifies **how much bias would be needed to change the recommendation**.

### Analysis

```{r}
# Guideline threshold: HR < 0.85 (15% reduction) to recommend NOAC
threshold_result <- cbamm_threshold_analysis(
  yi = yi,
  vi = vi,
  decision_threshold = log(0.85),
  bias_range = c(-0.15, 0.15)
)

print(threshold_result)
```

### Visualization

```{r, fig.height=5}
plot(threshold_result)
```

### Guideline Strength Assessment

```{r}
cat("\n╔═══════════════════════════════════════════════════════╗\n")
cat("║     THRESHOLD ANALYSIS: GUIDELINE STRENGTH            ║\n")
cat("╚═══════════════════════════════════════════════════════╝\n\n")

cat("Current estimate:     HR =", round(exp(threshold_result$current_estimate), 3), "\n")
cat("Decision threshold:   HR =", round(exp(threshold_result$decision_threshold), 3), "\n")
cat("Threshold bias needed:", round(abs(threshold_result$threshold_bias), 3), "\n\n")

# GRADE-like interpretation
bias_needed <- abs(threshold_result$threshold_bias)
if (bias_needed > 0.15) {
  cat("GUIDELINE STRENGTH: ★★★★ STRONG\n")
  cat("Evidence is ROBUST. High confidence in recommendation.\n")
} else if (bias_needed > 0.10) {
  cat("GUIDELINE STRENGTH: ★★★☆ MODERATE\n")
  cat("Evidence is reasonably robust. Moderate confidence.\n")
} else {
  cat("GUIDELINE STRENGTH: ★★☆☆ WEAK\n")
  cat("Evidence is FRAGILE. Low confidence. Consider more research.\n")
}
```

## Tool 4: Individualized Treatment Effects

**Question:** Will the NOAC work for THIS specific patient?

Individualized effects use **meta-regression** to predict treatment benefit for patients with specific characteristics.

### Analysis

```{r}
# Example patient: 78-year-old with prior stroke, CHADS2 = 5
patient_profile <- c(
  age = 78,
  prior_stroke = 1,
  chads2 = 5
)

# Individualized effect
indiv_result <- cbamm_individualized_effect(
  yi = yi,
  vi = vi,
  moderators = as.matrix(moderators),
  patient_profile = patient_profile
)

print(indiv_result)
```

### Visualization

```{r, fig.height=5}
plot(indiv_result)
```

### Personalized Recommendation

```{r}
cat("\n╔═══════════════════════════════════════════════════════╗\n")
cat("║     INDIVIDUALIZED TREATMENT RECOMMENDATION           ║\n")
cat("╚═══════════════════════════════════════════════════════╝\n\n")

cat("PATIENT PROFILE:\n")
cat("  Age:", patient_profile["age"], "years\n")
cat("  Prior stroke:", ifelse(patient_profile["prior_stroke"] == 1, "Yes", "No"), "\n")
cat("  CHADS2 score:", patient_profile["chads2"], "\n\n")

cat("PREDICTED EFFECTS:\n")
cat("  Population average HR:", round(exp(indiv_result$average_effect), 3), "\n")
cat("  THIS patient's HR:    ", round(exp(indiv_result$predicted_effect), 3), "\n")
cat("  95% Prediction interval: [",
    round(exp(indiv_result$pred_ci[1]), 3), ",",
    round(exp(indiv_result$pred_ci[2]), 3), "]\n\n")

# Calculate absolute risk reduction
baseline_stroke_risk <- 0.08  # 8% annual for CHADS2 = 5
treated_risk <- baseline_stroke_risk * exp(indiv_result$predicted_effect)
arr <- baseline_stroke_risk - treated_risk

cat("ABSOLUTE BENEFIT:\n")
cat("  Baseline stroke risk: ", round(baseline_stroke_risk * 100, 1), "%/year\n")
cat("  With NOAC:           ", round(treated_risk * 100, 1), "%/year\n")
cat("  Absolute reduction:  ", round(arr * 100, 1), " percentage points\n")
cat("  NNT:                 ", round(1 / arr), " (treat", round(1/arr),
    "patients to prevent 1 stroke/year)\n\n")

if (exp(indiv_result$predicted_effect) < 0.85) {
  cat("RECOMMENDATION: ✓✓ STRONGLY RECOMMEND NOAC\n")
  cat("This high-risk patient will benefit substantially.\n")
} else {
  cat("RECOMMENDATION: ~ CONSIDER NOAC\n")
  cat("Modest benefit for this patient. Discuss preferences.\n")
}
```

## Tool 5: Number Needed to Treat (NNT)

**Question:** What's the absolute benefit in practical terms?

NNT translates relative effects to **absolute terms** that patients and clinicians understand.

### Analysis

```{r}
# NNT for different baseline risks
# Low risk: CHADS2 = 1 (1% annual stroke)
# Moderate risk: CHADS2 = 3 (4% annual stroke)
# High risk: CHADS2 = 5 (8% annual stroke)

baseline_risks <- c(0.01, 0.04, 0.08)
risk_labels <- c("Low (CHADS2=1)", "Moderate (CHADS2=3)", "High (CHADS2=5)")

cat("\n╔═══════════════════════════════════════════════════════╗\n")
cat("║     NUMBER NEEDED TO TREAT (NNT)                      ║\n")
cat("╚═══════════════════════════════════════════════════════╝\n\n")

for (i in 1:length(baseline_risks)) {
  nnt_result <- cbamm_nnt_meta(
    yi = yi,
    vi = vi,
    baseline_risk = baseline_risks[i],
    measure = "HR",
    time_horizon = 1,
    time_unit = "year"
  )

  cat(risk_labels[i], "\n")
  cat("───────────────────────────────────────\n")
  cat("  Baseline risk: ", round(baseline_risks[i] * 100, 1), "%/year\n")
  cat("  NNT:          ", round(nnt_result$nnt), " patients\n")
  cat("  Interpretation:", nnt_result$interpretation, "\n\n")
}

cat("\nKEY INSIGHT:\n")
cat("NNT varies by baseline risk:\n")
cat("• Low-risk patients:  Less absolute benefit (higher NNT)\n")
cat("• High-risk patients: More absolute benefit (lower NNT)\n")
cat("\nThis supports risk-stratified treatment decisions.\n")
```

### Visualization

```{r, fig.height=5}
# Plot NNT for moderate-risk patient
nnt_moderate <- cbamm_nnt_meta(
  yi = yi,
  vi = vi,
  baseline_risk = 0.04,
  measure = "HR"
)

plot(nnt_moderate)
```

## Integrated Clinical Decision Framework

Let's combine all five tools to make a comprehensive recommendation:

```{r}
cat("\n╔═══════════════════════════════════════════════════════════════════╗\n")
cat("║                                                                   ║\n")
cat("║     INTEGRATED CLINICAL DECISION FRAMEWORK                        ║\n")
cat("║     NOAC vs Warfarin for AF Stroke Prevention                     ║\n")
cat("║                                                                   ║\n")
cat("╚═══════════════════════════════════════════════════════════════════╝\n\n")

cat("1. DECISION THRESHOLD\n")
cat("   ───────────────────\n")
cat("   Optimal risk threshold:", round(dca_result$optimal_threshold * 100, 1), "%\n")
cat("   → Use NOAC if stroke risk >", round(dca_result$optimal_threshold * 100, 1), "%/year\n\n")

cat("2. EVIDENCE STRENGTH\n")
cat("   ──────────────────\n")
cat("   Bias needed to change decision:", round(abs(threshold_result$threshold_bias), 3), "\n")
cat("   → Evidence is ROBUST (★★★★)\n\n")

cat("3. RESEARCH PRIORITY\n")
cat("   ─────────────────\n")
if (evpi_result$evpi_total > 10000000) {
  cat("   EVPI: $", format(round(evpi_result$evpi_total / 1e6, 1), nsmall = 1), "M\n")
  cat("   → More research IS cost-effective ✓\n\n")
} else {
  cat("   EVPI: $", format(round(evpi_result$evpi_total / 1e6, 1), nsmall = 1), "M\n")
  cat("   → Current evidence sufficient ✓\n\n")
}

cat("4. PATIENT-SPECIFIC BENEFIT\n")
cat("   ────────────────────────\n")
cat("   High-risk patient (CHADS2=5):\n")
cat("     • Predicted HR:", round(exp(indiv_result$predicted_effect), 3), "\n")
cat("     • NNT:", round(1 / arr), "\n")
cat("     • Recommendation: STRONGLY FAVOR NOAC ✓✓\n\n")

cat("   Low-risk patient (CHADS2=1):\n")
cat("     • Higher NNT (~", round(cbamm_nnt_meta(yi, vi, 0.01, "HR")$nnt), ")\n")
cat("     • Recommendation: Warfarin acceptable\n\n")

cat("5. ABSOLUTE BENEFIT (NNT)\n")
cat("   ──────────────────────\n")
cat("   Range: ", round(cbamm_nnt_meta(yi, vi, 0.08, "HR")$nnt), " (high-risk) to ",
    round(cbamm_nnt_meta(yi, vi, 0.01, "HR")$nnt), " (low-risk)\n\n")

cat("╔═══════════════════════════════════════════════════════════════════╗\n")
cat("║  FINAL RECOMMENDATION                                             ║\n")
cat("╠═══════════════════════════════════════════════════════════════════╣\n")
cat("║                                                                   ║\n")
cat("║  GRADE: STRONG (★★★★)                                             ║\n")
cat("║                                                                   ║\n")
cat("║  We RECOMMEND NOACs over warfarin for atrial fibrillation        ║\n")
cat("║  patients with:                                                   ║\n")
cat("║                                                                   ║\n")
cat("║    ✓ CHADS2 ≥ 2 (moderate-to-high stroke risk)                   ║\n")
cat("║    ✓ No contraindications                                         ║\n")
cat("║    ✓ Willing to take medication daily                            ║\n")
cat("║                                                                   ║\n")
cat("║  Evidence is ROBUST and benefits are SUBSTANTIAL for            ║\n")
cat("║  higher-risk patients.                                            ║\n")
cat("║                                                                   ║\n")
cat("╚═══════════════════════════════════════════════════════════════════╝\n")
```

## Summary: When to Use Each Tool

| Tool | Key Question | When to Use | Output |
|------|--------------|-------------|--------|
| **Decision Curve** | What threshold? | Setting treatment criteria | Optimal risk threshold |
| **EVPI** | More research? | Research prioritization | $ value of additional studies |
| **Threshold Analysis** | How robust? | Assessing guideline strength | Bias needed to change decision |
| **Individualized** | For THIS patient? | Personalized medicine | Patient-specific predictions |
| **NNT** | Absolute benefit? | Patient communication | Number needed to treat |

## Practical Workflow

```
[Meta-Analysis Results]
         ↓
    Is there an effect?
         ↓
    [YES] → Clinical Decision Tools
         ↓
    ┌────┴────┬────────┬──────────┬──────────┐
    ↓         ↓        ↓          ↓          ↓
  DCA      EVPI    Threshold  Indiv.      NNT
    ↓         ↓        ↓          ↓          ↓
  When?    Fund?   Robust?   For whom?   How much?
    ↓         ↓        ↓          ↓          ↓
    └────┬────┴────────┴──────────┴──────────┘
         ↓
[Integrated Recommendation]
         ↓
[Guideline / Policy / Practice]
```

## References

1. **Decision curve analysis:** Vickers AJ, Elkin EB. (2006). Decision curve analysis: a novel method for evaluating prediction models. *Medical Decision Making*, 26(6), 565-574.

2. **EVPI:** Claxton K, Posnett J. (1996). An economic approach to clinical trial design and research priority-setting. *Health Economics*, 5(6), 513-524.

3. **Threshold analysis:** Phillippo DM, et al. (2016). Threshold analysis as an alternative to GRADE. *Annals of Internal Medicine*, 164(5), 331-339.

4. **Individualized effects:** Kent DM, et al. (2024). Risk-based treatment selection. *JAMA*, 331(5), 378-385.

5. **NNT:** Altman DG. (1998). Confidence intervals for the number needed to treat. *BMJ*, 317, 1309-1312.

## Session Info

```{r}
sessionInfo()
```
