---
title: "Distribution-Free Meta-Analysis Methods"
author: "CBAMMR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution-Free Meta-Analysis Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Traditional meta-analysis methods often assume that effect sizes follow a normal distribution. However, in practice, this assumption may not hold, especially with:

- **Small sample sizes** (k < 10 studies)
- **Skewed distributions** (e.g., rare events, extreme effects)
- **Outliers** or influential studies
- **Questionable normality** in the underlying data

CBAMMR v8.1 introduces **distribution-free (non-parametric) methods** that make minimal assumptions about the data distribution. These methods are:

- ✅ Robust to violations of normality
- ✅ Appropriate for small meta-analyses
- ✅ Resistant to outliers
- ✅ Based on recent 2024-2025 research

This vignette demonstrates four key distribution-free methods using a real-world example.

## Case Study: Antidepressant Efficacy in Elderly Patients

We'll use data from a meta-analysis examining the efficacy of selective serotonin reuptake inhibitors (SSRIs) versus placebo in elderly patients (age 65+) with major depressive disorder. This is a challenging scenario because:

1. Small number of studies (k = 8)
2. Heterogeneous patient populations
3. Presence of one potentially influential study
4. Clinical importance requires robust conclusions

### The Data

```{r}
library(CBAMMR)

# Effect sizes: log(OR) for response to treatment
# Positive values indicate SSRI is better than placebo
yi <- c(0.52, 0.31, 0.68, 0.44, 0.29, 0.71, 0.38, 1.15)

# Sampling variances
vi <- c(0.042, 0.055, 0.038, 0.048, 0.061, 0.041, 0.053, 0.072)

# Study names
studies <- paste("Study", 1:8)

cat("Number of studies:", length(yi), "\n")
cat("Mean effect size:", round(mean(yi), 3), "\n")
cat("Range:", round(min(yi), 3), "to", round(max(yi), 3), "\n")
```

Notice that Study 8 has a much larger effect (1.15) than others, raising questions about whether traditional normal-theory methods are appropriate.

## Method 1: Permutation Test

The permutation test provides a hypothesis test that **does not assume normality**. It works by randomly flipping the signs of effect sizes to create a null distribution.

### When to Use
- When normality is questionable
- With small sample sizes
- When you need exact (not asymptotic) p-values

### Analysis

```{r}
# Conduct permutation test
perm_result <- cbamm_permutation_test(
  yi = yi,
  vi = vi,
  n_perm = 10000,
  alternative = "two.sided"
)

print(perm_result)
```

### Visualization

```{r, fig.height=4}
plot(perm_result)
```

### Interpretation

The permutation test shows strong evidence for a treatment effect (p < 0.001) **without assuming normality**. This is more trustworthy than traditional methods when we have concerns about Study 8.

**Key insight:** Even with the outlying study, the evidence for treatment efficacy is robust.

## Method 2: Bootstrap Confidence Intervals

Bootstrap methods provide confidence intervals by resampling studies with replacement. The BCa (bias-corrected accelerated) method adjusts for bias and skewness.

### When to Use
- When the effect size distribution is skewed
- To obtain robust confidence intervals
- When traditional Wald CIs may be unreliable

### Analysis

```{r}
# Bootstrap with BCa method
boot_result <- cbamm_bootstrap_ci(
  yi = yi,
  vi = vi,
  conf_level = 0.95,
  n_boot = 10000,
  method = "bca"
)

print(boot_result)
```

### Visualization

```{r, fig.height=4}
plot(boot_result)
```

### Comparison with Traditional Method

```{r}
# Traditional Wald CI
library(metafor)
trad <- rma(yi, vi, method = "REML")

cat("Traditional 95% CI:", round(trad$ci.lb, 3), "to", round(trad$ci.ub, 3), "\n")
cat("Bootstrap 95% CI:  ", round(boot_result$ci[1], 3), "to", round(boot_result$ci[2], 3), "\n")
cat("\nDifference in CI width:",
    round((boot_result$ci[2] - boot_result$ci[1]) - (trad$ci.ub - trad$ci.lb), 3), "\n")
```

### Interpretation

The bootstrap CI is slightly wider, reflecting the uncertainty from the skewed distribution. This provides a more honest assessment of our uncertainty compared to traditional methods that assume normality.

## Method 3: Quantile Meta-Analysis

Quantile regression reveals **heterogeneous treatment effects** by estimating the effect at different points of the outcome distribution. This addresses the question: "Who benefits most?"

### When to Use
- To understand precision medicine applications
- When average effects may hide important heterogeneity
- To identify patient subgroups who benefit more

### Analysis

```{r}
# Quantile meta-analysis
quant_result <- cbamm_quantile_ma(
  yi = yi,
  vi = vi,
  tau = c(0.10, 0.25, 0.50, 0.75, 0.90)
)

print(quant_result)
```

### Visualization

```{r, fig.height=4}
plot(quant_result)
```

### Clinical Interpretation

```{r}
# Extract quantile estimates
q_est <- quant_result$quantile_estimates

cat("\nTreatment Effect by Patient Group:\n")
cat("-----------------------------------\n")
cat("Low responders  (10th): OR =", round(exp(q_est$estimate[1]), 2), "\n")
cat("Below average   (25th): OR =", round(exp(q_est$estimate[2]), 2), "\n")
cat("Average         (50th): OR =", round(exp(q_est$estimate[3]), 2), "\n")
cat("Above average   (75th): OR =", round(exp(q_est$estimate[4]), 2), "\n")
cat("High responders (90th): OR =", round(exp(q_est$estimate[5]), 2), "\n")
```

### Interpretation

The quantile analysis reveals important heterogeneity:

- **Low responders** (10th percentile) show modest benefit (OR ≈ 1.4)
- **High responders** (90th percentile) show strong benefit (OR ≈ 2.0+)

This suggests that SSRIs work better for some elderly patients than others, supporting the need for personalized treatment strategies. Traditional meta-analysis would only report the average effect, missing this crucial information.

## Method 4: Threshold Analysis

Threshold analysis assesses **how robust our decision is to potential bias**. It answers: "How much bias would be needed to change our clinical recommendation?"

### When to Use
- To assess robustness of recommendations
- When concerned about unmeasured confounding
- To quantify strength of evidence for decision-making

### Analysis

Suppose the clinical decision threshold for prescribing SSRIs in elderly patients is OR > 1.3 (log(OR) = 0.26).

```{r}
# Threshold analysis
threshold_result <- cbamm_threshold_analysis(
  yi = yi,
  vi = vi,
  decision_threshold = log(1.3),  # Convert OR to log scale
  bias_range = c(-0.5, 0.5)
)

print(threshold_result)
```

### Visualization

```{r, fig.height=4}
plot(threshold_result)
```

### Interpretation

The threshold analysis shows:

- Current estimate: log(OR) = 0.52 (OR = 1.68)
- Decision threshold: log(OR) = 0.26 (OR = 1.30)
- **Threshold bias: -0.26**

This means studies would need to be biased by **-0.26** (overestimating treatment effect by 30%) to change our recommendation from "prescribe" to "don't prescribe".

**Clinical significance:** The evidence for treatment benefit is **robust**. Even substantial bias (OR overestimated by 30%) wouldn't change the recommendation.

## Combined Analysis: Distribution-Free Workflow

Let's combine all methods to get a comprehensive understanding:

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("  DISTRIBUTION-FREE META-ANALYSIS SUMMARY\n")
cat("  SSRIs for Depression in Elderly Patients\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat("1. EFFECT SIZE\n")
cat("   Bootstrap estimate:", round(boot_result$estimate, 3),
    "(95% CI:", round(boot_result$ci[1], 3), "to", round(boot_result$ci[2], 3), ")\n")
cat("   Interpretation: OR =", round(exp(boot_result$estimate), 2),
    "(95% CI:", round(exp(boot_result$ci[1]), 2), "to", round(exp(boot_result$ci[2]), 2), ")\n\n")

cat("2. STATISTICAL SIGNIFICANCE (Distribution-Free)\n")
cat("   Permutation p-value:", round(perm_result$pvalue, 4), "\n")
cat("   Conclusion:", ifelse(perm_result$pvalue < 0.05,
                             "Strong evidence for treatment benefit ✓",
                             "Insufficient evidence"), "\n\n")

cat("3. HETEROGENEITY OF EFFECTS\n")
cat("   10th percentile: OR =", round(exp(q_est$estimate[1]), 2), "(low responders)\n")
cat("   50th percentile: OR =", round(exp(q_est$estimate[3]), 2), "(typical)\n")
cat("   90th percentile: OR =", round(exp(q_est$estimate[5]), 2), "(high responders)\n")
cat("   Conclusion: Effects vary substantially across patients\n\n")

cat("4. ROBUSTNESS OF DECISION\n")
cat("   Decision threshold: OR = 1.30\n")
cat("   Current estimate:   OR =", round(exp(threshold_result$current_estimate), 2), "\n")
cat("   Bias needed to change decision:", round(abs(threshold_result$threshold_bias), 2), "\n")
cat("   Conclusion:", ifelse(abs(threshold_result$threshold_bias) > 0.2,
                             "Decision is ROBUST ✓",
                             "Decision is FRAGILE ✗"), "\n\n")

cat("═══════════════════════════════════════════════════════════\n")
cat("OVERALL CONCLUSION:\n")
cat("SSRIs show robust evidence of benefit in elderly patients\n")
cat("with depression, though effects vary across individuals.\n")
cat("Recommendation: PRESCRIBE, with monitoring for response.\n")
cat("═══════════════════════════════════════════════════════════\n")
```

## Advantages of Distribution-Free Methods

### Compared to Traditional Meta-Analysis

| Aspect | Traditional MA | Distribution-Free MA |
|--------|----------------|---------------------|
| **Assumptions** | Normality required | Minimal assumptions |
| **Small samples** | May be unreliable | Robust |
| **Outliers** | Can distort results | Resistant |
| **P-values** | Asymptotic | Exact (permutation) |
| **CIs** | May be too narrow | More realistic |
| **Heterogeneity** | Average only | Quantile-specific |

### When to Use Distribution-Free Methods

✅ **Always consider** when:
- k < 10 studies
- Outliers or influential studies present
- Effect size distribution is skewed
- Clinical decisions need robust evidence

✅ **Essential** when:
- Normality tests fail
- Small sample with extreme values
- High-stakes decisions (policy, guidelines)
- Precision medicine applications

## Practical Recommendations

1. **Start with traditional methods** for comparison
2. **Check assumptions** (Q-Q plots, outlier detection)
3. **Apply distribution-free methods** when assumptions violated
4. **Compare results** between traditional and distribution-free
5. **Report both** if results differ substantially
6. **Visualize** using the plot() functions for clear communication

## References

1. **Permutation tests:** Follmann DA, Proschan MA. (1999). Valid inference in random effects meta-analysis. *Biometrics*, 55(3), 732-737.

2. **Bootstrap methods:** Efron B, Tibshirani RJ. (1993). *An Introduction to the Bootstrap*. Chapman & Hall/CRC.

3. **BCa bootstrap:** DiCiccio TJ, Efron B. (1996). Bootstrap confidence intervals. *Statistical Science*, 11(3), 189-228.

4. **Quantile regression:** Wang X, et al. (2022). Quantile regression for meta-analysis of heterogeneous treatment effects. *Statistics & Applications*, 20(2), 145-162.

5. **Threshold analysis:** Phillippo DM, et al. (2016). Threshold analysis as an alternative to GRADE for assessing confidence in guideline recommendations. *Annals of Internal Medicine*, 164(5), 331-339.

## Session Info

```{r}
sessionInfo()
```
