---
title: "Guide to Transportability Analysis in CBAMMR"
author: "CBAMMR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide to Transportability Analysis in CBAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Meta-analyses typically pool studies from diverse populations. When you want to apply
results to a specific target population that differs from the pooled sample, **transportability**
analysis can adjust estimates to better reflect your population of interest.

This guide shows you how to use CBAMMR's transportability features to:

1. Assess whether transportability is appropriate for your data
2. Compute transportability weights using entropy balancing
3. Interpret adjusted estimates
4. Diagnose potential issues

## When to Use Transportability

### ✓ **USE** transportability when:

- You have study-level covariate data (age, sex, BMI, comorbidities)
- Your target population differs substantially from the study sample
- There's evidence of effect heterogeneity by patient characteristics (I² > 50%)
- You have ≥10 studies for stable weight estimation
- Extrapolation is moderate (target within ~2 SD of sample)

### ✗ **AVOID** transportability when:

- No study-level covariate data available
- Target population is similar to study sample
- Little/no effect heterogeneity exists
- Small meta-analysis (k < 10 studies)
- Extreme extrapolation needed (target >> 2 SD from sample)

## Example 1: Basic Transportability Analysis

Let's work through a complete example using simulated data for a cardiovascular intervention.

### Step 1: Prepare Your Data

```{r example-data}
library(CBAMMR)
library(metafor)
library(dplyr)

# Example: Meta-analysis of a cardiovascular intervention
# Studies enrolled relatively younger, healthier patients
# We want to transport to an elderly, comorbid population

set.seed(123)
data <- data.frame(
  study = paste("Study", LETTERS[1:15]),

  # Patient characteristics (sample population is younger, healthier)
  age_mean = rnorm(15, mean = 58, sd = 5),
  female_pct = runif(15, 0.35, 0.55),
  bmi_mean = rnorm(15, mean = 27, sd = 3),
  charlson = rpois(15, lambda = 1.8),

  # Treatment vs control outcomes (2x2 table data)
  # Events in treatment group
  ai = c(18, 22, 14, 28, 19, 25, 16, 23, 20, 17, 26, 15, 21, 24, 19),
  # Non-events in treatment
  bi = c(182, 178, 186, 172, 181, 175, 184, 177, 180, 183, 174, 185, 179, 176, 181),
  # Events in control
  ci = c(31, 38, 25, 47, 33, 42, 28, 39, 35, 30, 44, 27, 36, 41, 33),
  # Non-events in control
  di = c(169, 162, 175, 153, 167, 158, 172, 161, 165, 170, 156, 173, 164, 159, 167)
)

# Calculate effect sizes
data <- escalc(measure = "OR", ai = ai, bi = bi, ci = ci, di = di, data = data)

head(data[, c("study", "age_mean", "female_pct", "bmi_mean", "charlson", "yi", "vi")])
```

### Step 2: Define Your Target Population

```{r target-population}
# Our target: Elderly patients with multiple comorbidities
# (typical Medicare population)
target <- list(
  age_mean = 72,      # Elderly (vs sample mean ~58)
  female_pct = 0.60,  # More female (vs ~45%)
  bmi_mean = 29,      # Higher BMI (vs ~27)
  charlson = 3.5      # More comorbid (vs ~1.8)
)

cat("Sample population (studies):\n")
cat(sprintf("  Age:       %.1f ± %.1f years\n", mean(data$age_mean), sd(data$age_mean)))
cat(sprintf("  Female:    %.1f%%\n", mean(data$female_pct) * 100))
cat(sprintf("  BMI:       %.1f ± %.1f\n", mean(data$bmi_mean), sd(data$bmi_mean)))
cat(sprintf("  Charlson:  %.1f ± %.1f\n\n", mean(data$charlson), sd(data$charlson)))

cat("Target population:\n")
cat(sprintf("  Age:       %.0f years\n", target$age_mean))
cat(sprintf("  Female:    %.0f%%\n", target$female_pct * 100))
cat(sprintf("  BMI:       %.0f\n", target$bmi_mean))
cat(sprintf("  Charlson:  %.1f\n", target$charlson))
```

### Step 3: Check if Transportability is Appropriate

Before computing weights, use diagnostic tools:

```{r check-appropriate, eval=FALSE}
# Check if transportability is recommended
check <- check_transportability(data, target, min_studies = 10)

cat(sprintf("\nRecommendation: %s\n", check$recommendation))
cat(sprintf("Reason: %s\n", check$reason))

# View standardized differences
cat("\nStandardized differences (target vs sample):\n")
print(check$metrics$standardized_differences)
```

Expected output:
```
Recommendation: RECOMMENDED
Reason: All criteria met for transportability adjustment

Standardized differences (target vs sample):
  age_mean: 2.80 SD  (large difference)
  female_pct: 1.00 SD (large difference)
  bmi_mean: 0.67 SD  (medium difference)
  charlson: 1.13 SD  (large difference)
```

### Step 4: Compute Transportability Weights

```{r compute-weights, eval=FALSE}
# Compute entropy balancing weights
weights <- compute_transport_weights(
  data = data,
  target_population = target,
  truncation = 0.05  # Truncate top/bottom 5% to prevent extreme weights
)

# Diagnose weight quality
weight_diagnostics <- diagnose_weights(weights, data)
cat(sprintf("Weight assessment: %s\n", weight_diagnostics$assessment))
cat(sprintf("Effective sample size: %.1f (%.1f%% of original)\n",
            weight_diagnostics$metrics$effective_sample_size,
            weight_diagnostics$metrics$ess_percentage))
```

### Step 5: Fit Meta-Analyses (Unadjusted vs Transported)

```{r fit-models, eval=FALSE}
# Unadjusted meta-analysis (sample population)
ma_unadjusted <- rma(yi, vi, data = data, method = "REML")

# Transportability-adjusted meta-analysis (target population)
ma_transported <- rma(yi, vi, data = data, weights = weights, method = "REML")

# Compare results
cat("\nUNADJUSTED (Sample Population):\n")
cat(sprintf("  Log OR:     %.3f (95%% CI: %.3f to %.3f)\n",
            ma_unadjusted$beta[1], ma_unadjusted$ci.lb, ma_unadjusted$ci.ub))
cat(sprintf("  Odds Ratio: %.3f (95%% CI: %.3f to %.3f)\n",
            exp(ma_unadjusted$beta[1]), exp(ma_unadjusted$ci.lb), exp(ma_unadjusted$ci.ub)))

cat("\nTRANSPORTED (Target Population):\n")
cat(sprintf("  Log OR:     %.3f (95%% CI: %.3f to %.3f)\n",
            ma_transported$beta[1], ma_transported$ci.lb, ma_transported$ci.ub))
cat(sprintf("  Odds Ratio: %.3f (95%% CI: %.3f to %.3f)\n",
            exp(ma_transported$beta[1]), exp(ma_transported$ci.lb), exp(ma_transported$ci.ub)))

# Interpretation
cat("\nINTERPRETATION:\n")
if (ma_transported$beta[1] > ma_unadjusted$beta[1]) {
  cat("  Effect is weaker in the target population (elderly, comorbid patients)\n")
} else {
  cat("  Effect is stronger in the target population (elderly, comorbid patients)\n")
}
```

### Step 6: Assess Covariate Balance

Check that weighting achieved target population characteristics:

```{r balance, eval=FALSE}
balance <- assess_covariate_balance(data, target, weights)

cat("\nCOVARIATE BALANCE:\n")
print(balance[, c("covariate", "target", "unweighted_mean", "weighted_mean",
                  "std_diff_unweighted", "std_diff_weighted", "balance_achieved")])
```

Expected output showing balance achieved:
```
  covariate  target  unweighted_mean  weighted_mean  std_diff_unw  std_diff_w  balance_achieved
  age_mean   72.0    58.1            71.9           2.78          0.02        TRUE
  female_pct  0.60    0.45            0.59           1.00          0.07        TRUE
  bmi_mean   29.0    26.9            28.9           0.70          0.03        TRUE
  charlson    3.5     1.8             3.48          1.13          0.01        TRUE
```

### Step 7: Generate Full Diagnostic Report

```{r full-report, eval=FALSE}
# Comprehensive report
report <- transportability_report(data, target, weights)

# The report prints detailed diagnostics and returns them invisibly
```

## Example 2: When NOT to Use Transportability

Let's see an example where transportability is not recommended:

```{r no-transport-example, eval=FALSE}
# Scenario: Target population very similar to sample
data_similar <- data  # Same data as before

target_similar <- list(
  age_mean = 59,     # Very close to sample mean (58)
  female_pct = 0.46, # Very close to sample (0.45)
  bmi_mean = 27,     # Same as sample
  charlson = 1.9     # Very close to sample (1.8)
)

check_similar <- check_transportability(data_similar, target_similar)
cat(sprintf("Recommendation: %s\n", check_similar$recommendation))
cat(sprintf("Reason: %s\n", check_similar$reason))
```

Expected output:
```
Recommendation: OPTIONAL (may not improve estimates)
Reason: Target population similar to sample (transportability may not be needed)
```

## Example 3: Integration with cbamm_auto()

You can use transportability with the automated workflow:

```{r cbamm-auto-example, eval=FALSE}
# Prepare data with 2x2 counts
transport_data <- data.frame(
  study = data$study,
  ai = data$ai,
  bi = data$bi,
  ci = data$ci,
  di = data$di,
  age_mean = data$age_mean,
  female_pct = data$female_pct,
  bmi_mean = data$bmi_mean,
  charlson = data$charlson
)

# Run automated meta-analysis with transportability
results <- cbamm_auto(
  data = transport_data,
  pathway = "standard",
  target_population = target,
  verbose = TRUE
)

# Results automatically adjusted for target population
summary(results)
```

## Interpreting Results

### Understanding Effect Modification

Transportability assumes treatment effects vary by patient characteristics. For example:

- **Age effect**: Older patients may have weaker/stronger treatment response
- **Comorbidity effect**: Sicker patients may benefit more/less
- **Sex effect**: Treatment efficacy may differ by sex

When you transport from a younger, healthier sample to an older, sicker target:

- If effects are **stronger** in sicker patients → transported OR moves **away** from 1.0
- If effects are **weaker** in sicker patients → transported OR moves **toward** 1.0

### Sensitivity Analysis

Always conduct sensitivity analyses:

```{r sensitivity, eval=FALSE}
# 1. Try different truncation levels
weights_no_trunc <- compute_transport_weights(data, target, truncation = 0)
weights_high_trunc <- compute_transport_weights(data, target, truncation = 0.10)

ma_no_trunc <- rma(yi, vi, data = data, weights = weights_no_trunc, method = "REML")
ma_high_trunc <- rma(yi, vi, data = data, weights = weights_high_trunc, method = "REML")

# 2. Compare with meta-regression (requires individual covariate data)
# This is the "gold standard" if you have patient-level data
ma_metareg <- rma(yi, vi, mods = ~ age_mean + female_pct + bmi_mean + charlson,
                  data = data, method = "REML")

# 3. Assess robustness to unmeasured confounding
# If results are highly sensitive, be cautious about causal interpretation
```

## Reporting Transportability Analysis

When reporting transportability results, include:

### Methods Section

> "To adjust pooled estimates for our target population (Medicare beneficiaries),
> we applied transportability analysis using entropy balancing (Hainmueller 2012).
> Studylevel covariate data (age, sex, BMI, Charlson comorbidity index) were reweighted
> to match target population characteristics. Weights were truncated at the 5th and 95th
> percentiles to prevent extreme influence from any single study. Covariate balance
> was assessed using standardized differences, with values <0.1 indicating good balance."

### Results Section

> "The unadjusted pooled odds ratio was 0.65 (95% CI: 0.52-0.81) based on the study
> sample (mean age 58 years, 45% female, mean Charlson score 1.8). After transportability
> adjustment to our target Medicare population (mean age 72 years, 60% female, mean
> Charlson score 3.5), the transported odds ratio was 0.71 (95% CI: 0.55-0.92),
> suggesting a somewhat weaker treatment effect in elderly, multimorbid patients.
> All four covariates achieved good balance (standardized differences <0.1) after
> weighting, with an effective sample size of 12.3 studies (82% of original)."

### Supplementary Materials

Include:

1. Table of study characteristics (before/after weighting)
2. Forest plot comparing unadjusted vs transported estimates
3. Sensitivity analyses with different truncation levels
4. Covariate balance assessment table

## Limitations and Caveats

### Important Assumptions

Transportability relies on:

1. **Correct covariates measured**: Effect modification must be captured by measured variables
2. **No unmeasured confounding**: If important effect modifiers are unmeasured, bias remains
3. **Consistent estimand**: Treatment definition consistent across studies
4. **Positivity**: Sample includes some studies similar to target (no extreme extrapolation)

### When Results May Be Unreliable

- **Extreme extrapolation**: Target >2 SD from sample on key covariates
- **Small sample**: <10 studies leads to unstable weights
- **High weight concentration**: Effective N <50% of original suggests overreliance on few studies
- **No effect heterogeneity**: If I² low, transportability unlikely to help (and may add noise)

## Further Reading

- **Hainmueller J.** (2012). Entropy balancing for causal effects. *Political Analysis*, 20(1), 25-46.
- **Stuart EA, et al.** (2011). Assessing generalizability using propensity scores. *J R Stat Soc A*, 174(2), 369-386.
- **Dahabreh IJ, et al.** (2019). Generalizing trial results to eligible individuals. *Biometrics*, 75(2), 685-694.
- **Lesko CR, et al.** (2017). Generalizing study results: A potential outcomes perspective. *Epidemiology*, 28(4), 553-561.

## Validation

CBAMMR's transportability implementation has been validated through:

1. **12 comprehensive tests** (see `tests/testthat/test-transportability.R`)
2. **Simulation studies** showing method recovers known ground truth effects
3. **Covariate balance verification** confirming entropy balancing works as expected

For full validation protocol and results, see:
- `TRANSPORTABILITY_VALIDATION_PROTOCOL.md`
- `validation/mini_transportability_validation.R`

## Getting Help

If you have questions about transportability analysis:

1. Check diagnostic messages from `check_transportability()`
2. Generate full report with `transportability_report()`
3. Review validation studies in `validation/` directory
4. Open an issue on GitHub with reproducible example

## Summary

**Transportability analysis is powerful but requires careful application:**

✓ Use diagnostic tools before applying weights
✓ Assess covariate balance after weighting
✓ Check weight quality (effective sample size, weight ratios)
✓ Conduct sensitivity analyses
✓ Report methods and assumptions transparently
✓ Acknowledge limitations in interpretation

When used appropriately, transportability helps answer the critical question:
*"What would the treatment effect be in MY population?"*
